A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-13 paper.tex
rough cut. still missing last discussion section and finite ts VSI results

\usepackage{amsmath}
\usepackage{paralist}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\Htilde}{H_\epsilon}
\newcommand{\Hdust}{H_\mathrm{d}}
\newcommand{\Hgas}{H_\mathrm{g}}
\newcommand{\hgas}{h_\mathrm{g}}
\newcommand{\tepsilon}{\tilde{\epsilon}}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\Hg}{H_\mathrm{g}}
\newcommand{\Hd}{H_\mathrm{d}} 
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{lin15}{LY15}
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

@@

\title{On the entropy of 
  isothermal dusty gas: application to protoplanetary disks}  
\author{Min-Kai Lin\altaffilmark{1} and Andrew N. Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  %Tightly-coupled dust particles
  % in protoplanetary disks almost follow the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
\begin{inparaenum}[1)]
\item
  identify the entropy of isothermal dusty gas; and 
\item 
  generalize previous, well-known gaseous instabilities in
  protoplanetary disks to include dust dynamics. 
\end{inparaenum}
This includes the axisymmetric stability of strictly isothermal dusty
gas, allowing us to constrain the sharpness of 
dust edges that can persist in protoplanetary disks. We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results are consistent with our definition of
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential applications of the dusty/adiabatic gas
correspondence to provide thermodynamical interpretations of dusty
instabilities in protoplanetary disks.  
\end{abstract}

@@

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in 
fact the usual hydrodynamic equations. This leads to a precise 
definition of the entropy of an isothermal dusty gas. We describe approximate 
equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty problem in various
limits. We use these results to constraint the sharpness of dust edges
in isothermal disks. In \S\ref{linear_problem} we formulate an
explicit stability calcultion to study the dusty VSI, for which we
present numerical results in \S\ref{results}. We summarize and discuss
future applications of the dusty/adiabatic gas equivalence in
\S\ref{summary}.

>>>>

\usepackage{amsmath}
\usepackage{paralist}
\usepackage{url}
\usepackage{bm}
%\usepackage{hyperref}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\Htilde}{H_\epsilon}
\newcommand{\Hdust}{H_\mathrm{d}}
\newcommand{\Hgas}{H_\mathrm{g}}
\newcommand{\hgas}{h_\mathrm{g}}
\newcommand{\tepsilon}{\tilde{\epsilon}}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\Hg}{H_\mathrm{g}}
\newcommand{\Hd}{H_\mathrm{d}} 
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{lin15}{LY15}
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

@@

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks:
  axisymmetric stability}  
\author{Min-Kai Lin\altaffilmark{1} and Andrew N. Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
%\begin{inparaenum}[1)]
%\item
%  identify the entropy of isothermal dusty gas; and 
%\item 
  generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%\end{inparaenum}
%This includes the axisymmetric stability of strictly isothermal dusty
%gas, allowing us to constrain the sharpness of 
%dust edges that can persist in protoplanetary disks. 
We show that strictly isothermal, non-self-gravitating, axisymmetric
disks perfectly coupled to dust are generally stable, however thin the
dust layer is.  %except possibily in the presence of sharp 
%edges in radial dependence of the dust-to-gas ratio. 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential thermodynamical interpretations of 
dust-gas drag dynamics, for example the streaming instability 
as a manifestation of pulsational instabilities in stars. 
\end{abstract}

@@

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in 
fact the usual hydrodynamic equations. This leads to a precise 
definition of the entropy of an isothermal dusty gas. We describe approximate 
equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty problem in various
limits. We use these results to constraint the sharpness of dust edges
in isothermal disks. In \S\ref{linear_problem} we formulate an
explicit stability calcultion to study the dusty VSI, for which we
present numerical results in \S\ref{results}. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

### 2016-08-13 result.tex
rough cut. still missing last discussion section and finite ts VSI results

When $\tstop=0$ we may construct exact dusty equilibra by specifying
the dust-to-gas ratio, as described in \S\ref{eqm}. We use this to
examine the effect of dust on classic VSI, defined to be driven by the
temperature gradient (\S\ref{dusty_vsi_int}). For simplicity we
consider constant input values of the midplane dust-to-gas ratio
$\epsilon_0$ and the characteristic dust thickness $\Hd$. These,
together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used in \citetalias{lin15}.

>>>>

%When $\tstop=0$ we may construct exact dusty equilibra by specifying
%the dust-to-gas ratio, as described in \S\ref{eqm}.

We calculate explicit solutions to the linearized
equations in the limit of perfectly coupled dust ($\tstop=0$), so that
the equilibria defined in \S\ref{eqm} are exact steady states. 
We use this to
examine the effect of dust on classic VSI, defined to be driven by the
temperature gradient (\S\ref{dusty_vsi_int}). For simplicity we
consider constant input values of the midplane dust-to-gas ratio
$\epsilon_0$ and the characteristic dust thickness $\Hd$. These,
together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used in \citetalias{lin15}.

Two boundary conditions are needed for the $\tstop=0$
problem. This is expected since this problem is exactly equivalent to
adiabatic hydrodynamics \citep[e.g.][]{lubow93}. For simplicity we
impose solid boundaries so that $\delta v_z(\pm\zmax)=0$.  
We solve the linearized equations as a generalized eigenvalue problem 
using a pseudo-spectral code adapted from \cite{lin15}. Amplitudes 
are expanded in Chebyshev polynomials up to order $N_z=512$. We check
results using Eq. \ref{vsi_check}.

### 2016-08-13 summary.tex
rough cut. still missing last discussion section and finite ts VSI results

\section{Summary and discussion}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. 
In the limit of perfect dust-gas coupling, the dust content of a fluid 
parcel $\rhod/\rhog$ is conserved. This is analogous to entropy being 
conserved following a adiabatic, pure gas. %s = s(\rhod/\rhog)
We explicitly show that for 
a locally isothermal, dusty gas that the evolutionary equation of 
$\rhod/\rhog$ may be recast into an effective energy equation identical
in form to that in standard adiabatic hydrodynamics. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
This allows us to define the buoyancy frequency of an isothermal 
dusty gas.

%growth rates not sig changed 
%v shear from thin dust layers are overwhelmed by

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}
  \frac{D P}{D t} = - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla \ln{K}
  + \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  s = \ln{\left[K(r,z)\left(1 - \tepsilon\right)^\Gamma\right]}.  
\end{align}
When $K$ is constant, the corresponding Solberg-Hoiland criteria for
axisymmetric stability may be obtained from that in standard adiabatic 
hydrodynamics \citep[e.g.][]{tassoul78} by
using this definition of entropy in
Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

@@

%\subsection{Implication for numerical simulations}

\subsection{Thermodynamic interpretation of dusty instabilities}

%\subsection{Caveats and future directions}
%gi

>>>>

%growth rates not sig changed 
%v shear from thin dust layers are overwhelmed by 
\section{Discussion}\label{discussion}

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} = - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla \ln{K}
  + \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  s = \ln{\left[K^\frac{1}{\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

@@

\subsection{Understanding dust-drag
  instabilities}\label{dust_as_thermo}

\section{Summary}\label{summary}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. 
In the limit of perfect dust-gas coupling, the dust content of a fluid 
parcel $\rhod/\rhog$ is conserved. This is analogous to entropy being 
conserved following a adiabatic, pure gas. %s = s(\rhod/\rhog)
We explicitly show that for 
a locally isothermal, dusty gas that the evolutionary equation of 
$\rhod/\rhog$ may be recast into an effective energy equation identical
in form to that in standard adiabatic hydrodynamics. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
This allows us to define the buoyancy frequency of an isothermal 
dusty gas.
